---
title: "Google Capstone Project"
author: "SenaTheAnalyst"
date: "2023-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Background 

The director of marketing of a fictional Company Cyclistic believes the company’s future success
depends on maximizing the number of annual memberships. The analytics team wants to understand how
casual riders and annual members use Cyclistic bikes differently. From these insights, the marketing team will design
a new marketing strategy to convert casual riders into annual members. Based on analysis, Cyclistic executives will
approve recommendations, so they must be backed up with compelling data insights and professional
data visualizations.

**Key stakeholders**:\
Lily Moreno: The director of marketing,\
Cyclistic executive team and\
Cyclistic marketing analytics team.


**Cyclistic**: A bike-share program that features more than 5,800 bicycles and 600 docking stations.
Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more
inclusive to people with disabilities and riders who can’t use a standard two-wheeled bike. The majority of
riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely
to ride for leisure, but about 30% use them to commute to work each day.

### Project Goal:

As a Junior Data Analysis, the goal is to determine how annual members and casual riders use
Cyclistic bikes differently with the aim of converting casual riders into annual member.

Based on insights from analysis, a new marketing strategy will be designed to convert casual riders to annual riders


### Tools Used for this project.

**Excel, SQL and R-Language**



### Prepare Phase

The data sets being used for the project are 12 months of trip data owned
by Motivate International Inc who have granted a non-exclusive,
royalty-free, limited, perpetual license to access, reproduce, analyze,
copy, modify, distribute in product or service and use the Data for
any lawful purpose ("License").

There are no issues of bias or credibility with these data sets. They are
Reliable, Original, Comprehensive, Current and Cited.

Upon initial inspection of each of the 12 csv data sets using ISBLANK
function in excel, i noticed that some of the observations are missing.
This i will address subsequently using RStudio.



## Installing packages

```{r}
r = getOption("repos")
r["CRAN"] = "http://cran.us.r-project.org"
options(repos = r)

install.packages("tidyverse",repos = "http://cran.us.r-project.org")
install.packages("janitor",repos = "http://cran.us.r-project.org")
install.packages("lubridate",repos = "http://cran.us.r-project.org")
install.packages("lifecycle",repos = "http://cran.us.r-project.org")
install.packages("pillar",repos = "http://cran.us.r-project.org")
install.packages("pillar") 
install.packages("dplyr")

```

## Loading packages

```{r}

library("janitor")
library("lubridate")
library("ggplot2")
library("readxl")
library("dplyr")
library("tidyverse")
library("tidyr")

```
## Loading Datasets

Loading the needed 12 csv data sets, 

Read CSV files (Data of trips by Cyclistic riders)


```{r}
df1  <- read.csv("202201-divvy-tripdata.csv")
df2  <- read.csv("202202-divvy-tripdata.csv")
df3  <- read.csv("202203-divvy-tripdata.csv")
df4  <- read.csv("202204-divvy-tripdata.csv")
df5  <- read.csv("202205-divvy-tripdata.csv")
df6  <- read.csv("202206-divvy-tripdata.csv")
df7  <- read.csv("202207-divvy-tripdata.csv")
df8  <- read.csv("202208-divvy-tripdata.csv")
df9  <- read.csv("202209-divvy-tripdata.csv")
df10 <- read.csv("202210-divvy-tripdata.csv")
df11 <- read.csv("202211-divvy-tripdata.csv")
df12 <- read.csv("202212-divvy-tripdata.csv")

```

## Binding all csv files to a common file for analysis

```{r}

bike_rides <- rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10,df11,df12)
head(bike_rides)
glimpse(bike_rides)


```

## Renaming some columns

```{r}
renamed_col <- bike_rides %>% 
  rename(start_time = started_at, end_time = ended_at,user_status = member_casual
         , bike_type = rideable_type)
glimpse(renamed_col)
head(renamed_col)

```

## Renaming some columns

```{r}
renamed_col <- bike_rides %>% 
  rename(start_time = started_at, end_time = ended_at,user_status = member_casual
         , bike_type = rideable_type)
glimpse(renamed_col)
head(renamed_col)

```

## Checking duration of trips(end time minus start time)

```{r}
trip_duration <-mutate(renamed_col, trip_duration = as.duration
                            (interval(ymd_hms(renamed_col$start_time),
                                      ymd_hms(renamed_col$end_time))))

```

## Adding another column to determine weekdays of the ride

```{r}
day_of_week <- trip_duration %>% mutate(day_of_week = weekdays(as.Date(start_time)))
View(day_of_week)


```

# Checking some variables (Consistency) in the columns with unique()

```{r}
unique(day_of_week$user_status)
unique(day_of_week$bike_type)

```

# checking and removing all (null) and (na) data for ease of data analysis

```{r}
is.null(day_of_week)
clean_trip_data <-na.omit(day_of_week) 
head(clean_trip_data)

```

## Analyzing data and Sharing results.


```{r}


user_status_count <-table(clean_trip_data$user_status)
user_status_count
pie(user_status_count)

pie(user_status_count,
    labels=as.character(user_status_count),
    main="Type of bike riders",
    col=c("blue","orange"),
    border="black",
    clockwise=TRUE
)


barplot(user_status_count,
        main="Type of bike riders",
        xlab="Bike riders",
        ylab="Count",
        border="blue",
        col="blue",
        density=20
 ) 

```

## From the above graphs it can conclude that the number of members (3344996) are more than casual riders (2316863)



# Now the average ride duration for both groups is determined

```{r}
avg_trip_duration <- clean_trip_data %>% 
  group_by(user_status) %>% 
  summarize(avg = mean(trip_duration)/60)
avg_trip_duration

View(clean_trip_data)

```

## On an average, casual riders ride longer (22 mins) than members (12.4 mins)



## Next determine day of the week when majority of rides are taken based on user status 

```{r}
day_of_ride <- clean_trip_data %>% 
  group_by(day_of_week , user_status) %>% 
  select(user_status, day_of_week)  %>% 
  summarize(number = table(day_of_week))

View(day_of_ride)

```

## In order to make the necessary extractions from and groupings, the table is exported to SQL.

## The command write.csv()  is used to create a data frame to be exported to SQL

```{r}
write.csv(day_of_ride, "day_of_week.csv")
```


## Import queried data for further analysis

**getwd()** : Is used to get the current working directory

casual_ride <- read.csv("C:/Users/SENANU/Documents/temp/Cyclistic_Analysis/Cyclsitic_Analysis/casual.csv")

member_ride <-read.csv("C:/Users/SENANU/Documents/temp/Cyclistic_Analysis/Cyclsitic_Analysis/members.csv")



```{r}
getwd() 

casual_ride <- read.csv("C:/Users/SENANU/Documents/temp/Cyclistic_Analysis/Cyclsitic_Analysis/casual.csv")

member_ride <-read.csv("C:/Users/SENANU/Documents/temp/Cyclistic_Analysis/Cyclsitic_Analysis/members.csv")


```

## Now sorting of imported data
## First casual riders

```{r}
sort_casual_ride <- casual_ride %>% 
  arrange(desc(number)) 

sort_casual_ride

```

## Next members riders

```{r}

sort_member_ride <- member_ride %>% 
  arrange(desc(number)) 

sort_member_ride

```

## The values are plotted to get a clearer picture of casual riders

```{r}
ggplot(data = sort_casual_ride, aes(x=day_of_week, y=number, color =day_of_week, fill= day_of_week)) 
  labs(title = 'Number of Casual Riders vs Days of the week', 
       caption = 'Analysis by SenaTheAnalyst')+ 
  geom_bar(stat = "identity")+ theme(legend.position ="none")



ggplot(data = sort_member_ride, aes(x=day_of_week, y=number, color =day_of_week, fill= day_of_week)) + 
  labs(title = 'Number of Member Riders vs Days of the week', 
       caption = 'Analysis by SenaTheAnalyst')+ 
  geom_bar(stat = "identity")+theme(legend.position="none")


```

### From the above graph, it is evident that majority of the casual riders, close to 60%, prefer riding over the weekend, Fridays, Saturdays and Sundays as against the member/registered riders that are almost evenly distributed across the weeks.
## About 60% of casual trips take place over the weekend.  





### To find out how the ridership varies by month based on user types, data is exported to SQL to make necessary extractions
### First the date format has to be chnaged to make t workable

```{r}
clean_trip_data$month_of_ride <- format(as.Date(clean_trip_data$end_time),"%m")

View(clean_trip_data)


```

### Converting numeric months to month_names

```{r}

clean_trip_data$name_of_months <- month.name[as.numeric(clean_trip_data$month_of_ride)]
View(clean_trip_data)

```

### Now finding out ridership by month

```{r}
ride_month <- clean_trip_data %>% 
  group_by(name_of_months,user_status) %>% 
  select(user_status,name_of_months) %>% 
  summarise(number= table(name_of_months))
ride_month

```

### Data is exported to SQL for analysis by first converting to csv file

```{r}
write.csv(ride_month, "ride_month.csv")
```

### Read data form SQL analysis

```{r}
casual_month_ride <- read.csv("C:/Users/SENANU/Documents/temp/Cyclistic_Analysis/Cyclsitic_Analysis/casual_month_of _year.csv")

member_month_ride <-read.csv("C:/Users/SENANU/Documents/temp/Cyclistic_Analysis/Cyclsitic_Analysis/members_month_of_year.csv")

casual_month_ride

member_month_ride

```

### The values are plotted to get a clearer picture of casual riders


```{r}

ggplot(data = casual_month_ride, aes(x=name_of_months, y=number, color =name_of_months, fill= name_of_months)) + 
  labs(title = 'Casul rides by the months of the year', 
       caption = 'Analysis by SenaTheAnalyst')+ 
  geom_bar(stat = "identity")+ theme(legend.position ="none")



ggplot(data = member_month_ride, aes(x=name_of_months, y=number, color =name_of_months, fill= name_of_months)) + 
  labs(title = 'Number of Member Riders vs Days of the week', 
       caption = 'Analysis by SenaTheAnalyst')+ 
  geom_bar(stat = "identity")+theme(legend.position="none")


```

### From the above graphs, it is evident that casual riders are mostly active from May to September which represents the end of spring to summer season.





## Highlights of analysis

* Casual riders ride an average of 22.0 minutes while members ride 12.4 on an average.
* There are more member riders (3344996) than casual riders(2316863).
* Casual rides bike usage turns to increase towards the end of the week (Thursday, Friday, Saturday & Sunday) and peaking on Saturdays.
* Members usage turns to be almost constant through the week but a slight increase between Tuesday to Thursday.
* Casual riders turn to purchase Cyclistic bikes during the summer months (June, July,August,September)
* Member riders also use bikes during the summer and ridership falls during the winter periods.



> Recommendations in order to convert casual riders to member riders.

1. Offer discounted packages for casual rides that  last longer that 20 minutes during the weekend.
2. Launch a weekend ride to win promotion for riders by duration of ride.
3. Offer membership exclusive incentives/packages for weekend rides.  
4. Offer discounted special summer packages to casual riders.  
 


